k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-03-14 paper.Rmd
shown here
introduction

Partitioning data sets via anticlustering has uses in many research fields including psychology [@brusco2019; @schaper2023; @papenberg2020], test assembly [@gierl2017], education [@krauss2013; @baker2002], artificial intelligence [@steghofer2013], machine learning [@mauri2023], network systems [@mohebi2022], and operations research [e.g. @gallego2013; @gliesch2021]. Anticlustering seeks for similarity between groups and heterogeneity within groups by maximizing a clustering objective criterion. Thus, anticlustering is the logical and mathematical opposite of classical clustering, which partitions data sets via minimization of the same criteria [@steinley2006; @brusco2019; @spath1986].

@@

\begin{align}
\vert C_k \vert = n_k, \; (k = 1, \ldots, K) \label{formalization:1} \\
\sum_{k = 1}^{K} n_k = N \label{formalization:2} \\
C_j \cap C_k = \emptyset (j, k = 1, \ldots, K), \; j \ne k\label{formalization:3} \\
\end{align}

@@

We assume that a data matrix $\mathbf{Z}$ is used as input for computing the anticlustering criterion on the basis of a partitioning $C$. $\mathbf{Z}$ is either a feature matrix $\mathbf{X}_{N \times M} = \{x_{ij}\}_{N \times M}$ where the $M$ columns contain numeric measurements on the $N$ objects, or a symmetric matrix $\mathbf{D}_{N \times N} = \{d_{ij}\}_{N \times N}$ of pairwise dissimilarity measurements. The partitioning is obtained by maximizing a clustering criterion $f_i(C, \mathbf{Z})$. For example, *k*-means anticlustering aims to maximize between-group similarity by maximizing the error sum of squares, i.e., the sum the squared distances between each object and its cluster center [@spath1986; @papenberg2024]:

@@

Maximizing the diversity simultaneously minimizes the sum of distances between objects not in the same group. Therefore, the diversity represents within-group heterogeneity as well as between-group similarity [@feo1990]. Being a sum instead of an average, however, it no longer adequately captures between-group similarity if group sizes are not equal.

The diversity is oftentimes obtained by first converting the data matrix $\mathbf{X}_{N \times M}$ into a matrix of pairwise Euclidean distances [@brusco2019; @papenberg2020; @papenberg2024; @gallego2013]. If $\mathbf{D}_{N \times N}$ is obtained by using the squared pairwise Euclidean distances instead, we note an important equivalence between the k-means criterion $f_1$ and the diversity $f_3$. Traditionally, the k-means criterion is denoted via a sum of squared distances between data points and cluster centroids. However, it can also be expressed using the pairwise squared Euclidean distances between data points (Brusco, 2006; Späth, 1980). If we let $\mathbf{D'}_{N \times N}$ = \{$d'_{ij}$\} represent the squared Euclidean distance, i.e., $d'_{ij} = \sum\limits_{m = 1}^{M}(x_{im} - x_{jm})^2$, the k-means criterion can be expressed as as

@@

$$
f_4^*(C, \mathbf{D'}_{N \times N}) = 
  \frac{N}{K} \sum\limits_{k = 1}^{K} \sum_{(i<j)\in C_k} d_{ij}
$$

In the context of maximization, the normalizing factor $\frac{N}{K}$ can be ignored because it does not depend on the partitioning $C$. Hence,

@@

For anticlustering, the criterion $f_4^*$ (i.e., the k-means criterion for equal-sized groups) reduces to the sum of the (squared) distances between elements in the same cluster, i.e., the standard diversity $f_3$. Therefore, for equal-sized groups, the diversity anticlustering criterion $f_3$ is equivalent to the k-means criterion $f_1$---if the diversity is computed via the squared Euclidean distance. We will use this insight later when we discuss applying algorithms for maximum diversity to the $k$-means and $k$-plus objectives. For arbitrary measures of dissimilarity (and not just the squared Euclidean distance), we denote $f_4$ as the *average diversity criterion*.

Unlike $f_1$, $f_2$, $f_3$, and $f_4$, the *dispersion* criterion does not represent between-group similarity, but is a pure measure of within-group heterogeneity. It is defined as the minimum dissimilarity across any two objects that are part of the same group [@brusco2019; @fernandez2013]:

@@

The dispersion is a measure of the worst-case pairwise dissimilarity. Maximizing the dispersion ensures that any two objects in the same group are as dissimilar from each other as possible.

@brusco2019 outlined the importance of simultaneously considering multiple criteria when addressing anticlustering problems. In particular, they argued that anticlustering algorithms should incorporate an objective of between-group similarity---such as the diversity---and pairwise within-group dissimilarity---i.e., the dispersion---at the same time. To this end, @brusco2019 presented a bicriterion algorithm that simultaneously maximizes the diversity $f_3$ and the dispersion $f_5$. We endorse Brusco et al.'s bicriterion approach for anticlustering, and advance it in several ways in the current paper. First, we allow that not only the diversity $f_3$ can be optimized alongside the dispersion, but instead we allow that the k-means, k-plus or the average diversity can be used as measures of between-group similarity (i.e., the criteria $f_1$, $f_2$ and $f_4$). Second, we allow that the dispersion can be optimized on the basis of a different dissimilarity matrix than the diversity (or a different objective of between-group similarity). This is useful if the criteria should be based on different information, or to induce custom cannot-link constraints, as we will show. As our most important contribution, we combine a novel exact algorithm for maximum dispersion with the BILS heuristic to obtain high diversity on top of an optimal dispersion. Our paper is organized as follows. In the next section, we review the bicriterion method by @brusco2019 in detail. After that, we provide the technical background for our extensions. We then provide several evaluations of our contributions: (a) we probe the practical feasibility of our optimal algorithm for maximum dispersion, (b) we compare different adaptations of the BILS with regard to their ability to providing high diversity on top of an optimal dispersion, and (c) we provide practical examples on using our new methods.

>>>>

Partitioning data sets via anticlustering has uses in many research fields including psychology [@brusco2019; @schaper2023], test assembly [@gierl2017], education [@krauss2013; @baker2002], artificial intelligence [@steghofer2013], machine learning [@mauri2023], network systems [@mohebi2022], and operations research [e.g. @gallego2013; @gliesch2021]. In psychological research, an important application is the assignment of stimuli to different experimental sets that are presented in within-subjects designs: The different sets should be as similar as possible with regard to variables that affect the responses, and presenting the same set repeatedly is usually prohibited due to carry-over effects. Anticlustering is a powerful tool for automatically solving such tasks, which are time-consuming and usually inadequately handled manually [@lahl2006; @lintz2021; @cutler1981]. The first introduction to anticlustering in a Psychological journal was provided by @brusco2019. @papenberg2020 presented the free and open source software package `anticlust` that implements many algorithms for anticlustering, thus providing researchers with a tool for solving a problem that frequently occurs in psychological research, but previously lacked an appropriate software solution.

@@

\begin{align}
\vert C_k \vert = n_k, \; (k = 1, \ldots, K) \label{formalization:1} \\
\sum_{k = 1}^{K} n_k = N \label{formalization:2} \\
C_j \cap C_k = \emptyset (j, k = 1, \ldots, K), \; j \ne k\label{formalization:3}
\end{align}

@@

Anticlustering seeks for similarity between groups and heterogeneity within groups by maximizing a clustering objective criterion. Thus, anticlustering is the logical and mathematical opposite of classical clustering, which partitions data sets via minimization of the same criteria [@steinley2006; @brusco2019; @spath1986]. We assume that a data matrix $\mathbf{Z}$ is used as input for computing the anticlustering criterion on the basis of a partitioning $C$. $\mathbf{Z}$ is either a feature matrix $\mathbf{X}_{N \times M} = \{x_{ij}\}_{N \times M}$ where the $M$ columns contain numeric measurements on the $N$ objects, or a symmetric matrix $\mathbf{D}_{N \times N} = \{d_{ij}\}_{N \times N}$ of pairwise dissimilarity measurements. The partitioning is obtained by maximizing a clustering criterion $f_i(C, \mathbf{Z})$. For example, *k*-means anticlustering aims to maximize between-group similarity by maximizing the error sum of squares, i.e., the sum the squared distances between each object and its cluster center [@spath1986; @steinley2006]:

@@

Maximizing the diversity $f_3$ simultaneously minimizes the sum of distances between objects not in the same group. Therefore, the diversity represents within-group heterogeneity as well as between-group similarity [@feo1990]. Being a sum instead of an average, however, it no longer adequately captures between-group similarity if group sizes are not equal [@papenberg2024].

The diversity is oftentimes computed by first converting a feature matrix $\mathbf{X}_{N \times M}$ into a matrix of pairwise Euclidean distances [@brusco2019; @papenberg2020; @papenberg2024; @gallego2013]. If $\mathbf{D}_{N \times N}$ is obtained by using the squared pairwise Euclidean distances instead, we note an important equivalence between the k-means criterion $f_1$ and the diversity $f_3$. Traditionally, the k-means criterion is denoted via a sum of squared distances between data points and cluster centroids. However, it can also be expressed using the pairwise squared Euclidean distances between data points (Brusco, 2006; Späth, 1980). If we let $\mathbf{D'}_{N \times N}$ = \{$d'_{ij}$\} represent the squared Euclidean distance, i.e., $d'_{ij} = \sum\limits_{m = 1}^{M}(x_{im} - x_{jm})^2$, the k-means criterion can be expressed as $f_4$:

@@

$$
f_4^*(C, \mathbf{D'}_{N \times N}) = 
  \frac{K}{N} \sum\limits_{k = 1}^{K} \sum_{(i<j)\in C_k} d_{ij}
$$

In the context of maximization, the factor $\frac{K}{N}$ can be ignored because it does not depend on the partitioning $C$.

@@

Therefore, the criterion $f_4^*$ (i.e., the k-means criterion for equal-sized groups) reduces to the standard diversity $f_3$---if the diversity is computed via the squared Euclidean distance. We will use this insight later when we discuss applying algorithms for maximum diversity to the $k$-means and $k$-plus objectives. For arbitrary measures of dissimilarity (and not just the squared Euclidean distance), we denote $f_4$ as the *average diversity* criterion.

Unlike $f_1$, $f_2$, $f_3$, and $f_4$, the *dispersion* criterion does not represent between-group similarity, but is a pure measure of within-group heterogeneity. It is defined as the minimum dissimilarity across any two objects that are part of the same group [@fernandez2013]:

@@

The dispersion is a measure of the worst-case pairwise dissimilarity [@brusco2019]. Maximizing the dispersion ensures that any two objects in the same group are as dissimilar from each other as possible, which is desireable when striving for high within-group heterogeneity.

@brusco2019 outlined the importance of simultaneously considering multiple criteria when addressing anticlustering problems. In particular, they argued that anticlustering algorithms should incorporate an objective of between-group similarity such as the diversity, and pairwise within-group dissimilarity, i.e., the dispersion. To this end, @brusco2019 presented a bicriterion algorithm (*BILS*) that simultaneously maximizes the diversity $f_3$ and the dispersion $f_5$ by approximating the Pareto efficient set of partitions according to both criteria. By investigating the Pareto set for many problem instances, they found that a considerable improvement in one of these criteria is often possible with little or even no sacrifice in the other. We endorse Brusco et al.'s bicriterion approach for anticlustering and extend it in several ways. First, we allow that not only the diversity $f_3$ can be optimized alongside the dispersion, but we also allow that the k-means, k-plus or the average diversity can be employed as measures of between-group similarity during the bicriterion optimization (i.e., the criteria $f_1$, $f_2$ and $f_4$). Second, we allow that the dispersion can be optimized on the basis of a different dissimilarity matrix than the diversity (or a different objective of between-group similarity). This is useful if the objectives should be based on different information, or, as we discuss below, to induce custom pairwise cannot-link constraints. As our most important contribution, we present a novel exact algorithm for maximum dispersion and investigate different algorithmic techniques for providing high diversity on top of it. All of the methods newly developed here as well as the original BILS algorithm are available via the free and open source `R` package `anticlust` [@papenberg2020; @R-anticlust].

Our paper is organized as follows. In the next section, we discuss algorithms for maximizing anticlustering criteria. In particular, we review the bicriterion algorithm by @brusco2019 in detail. After that, we provide the technical background for our own extensions of the bicriterion approach. We continue by evaluating these contributions: (a) we probe the practical feasibility of our optimal algorithm for maximum dispersion for varying $N$ and $K$; (b) we compare different algorithmic approaches in their ability to provide high diversity on top of an optimal dispersion; (c) we apply our new methods to real norming data and present functioning `R` code to reproduce the examples [@R-base].
